New numerical integration of orbits of irregular satellites and prediction of stellar occultations up to 2020

### 2016-03-07 Irregular-occ/Irregular-occ.tex
.

\title[Predictions of stellar occultations by irregular satellites up to 2020]{Stellar occultation predictions for 10 irregular satellites of giant planets plus Triton up to 2020}
\author[A. R. Gomes-J\'unior, M. Assafin, L. Beauvalet et al.]{A. R. Gomes-J\'unior$^{1}$\thanks{E-mail: altair08@astro.ufrj.br},
M. Assafin$^{1,\dag}$,
L. Beauvalet$^{2,3}$,
J. Desmars$^{4}$,
R. Vieira-Martins$^{1,2,5,\dag}$,\newauthor
J. I. B. Camargo$^{2,5}$,
B. E. Morgado$^{1,2}$
F. Braga-Ribas$^{2,6}$
\\
$^{1}$Observat\'orio do Valongo/UFRJ, Ladeira Pedro Ant\^onio 43,
CEP 20.080-090 Rio de Janeiro - RJ, Brazil\\
$^{2}$Observat\'orio Nacional/MCTI, R. General Jos\'e Cristino 77, CEP 20921-400 Rio de Janeiro - RJ, Brazil\\
$^{3}$Observatoire de Paris/SYRTE, 77 Avenue Denfert Rochereau 75014 Paris, France\\
$^{4}$Institut de m\'ecanique c\'eleste et de calcul des \'eph\'em\'erides - Observatoire de Paris, UMR 8028 du CNRS, 77 Av. Denfert-Rochereau,\\ 75014 Paris, France\\
$^{5}$Laborat\'orio Interinstitucional de e-Astronomia - LIneA, Rua Gal. Jos\'e Cristino 77, Rio de Janeiro, RJ 20921-400, Brazil\\
$^{6}$Federal University of Technology - Paran\'a (UTFPR / DAFIS), Rua Sete de Setembro, 3165, CEP 80230-901, Curitiba, PR, Brazil
}
\begin{document}
\newcommand{\noccs}{5457 } % total number of occultations

@@

\begin{abstract}
Due to their orbital configurations, it is common belief that the irregular satellites were probably captured by the giant planets in the early solar system. It is important to know their physical parameters, such as size, shape, albedo and composition, to trace back their true origin. The best ground-based technique to determine size and shape, and thus constrain the albedo and in a broader sense composition, is the observation of stellar occultations by these objects.
We aim to predict stellar occultations for the eight largest irregular satellites of Jupiter: Himalia, Elara, Pasiphae, Carme, Lysithea, Sinope, Ananke and Leda, and for the irregular satellites Phoebe of Saturn and Nereid of Neptune, and also for Triton.
We identified candidates to stellar occultations by the irregular satellites from the UCAC4 catalogue and from a catalogue of stars in the sky-path of Neptune obtained from observations made with the ESO2p2/WFI (2.2 m Max-Planck ESO telescope with the Wide Field Imager) instrument. These catalogues were crossed with the ephemeris of the satellites to identify stellar occultations. We used a new ephemeris based solely on the observations from \cite{GomesJunior2015} to generate predictions for the short-time future of the satellites of Jupiter. For Phoebe, we used an updated ephemeris and for Nereid and Triton we used recently published ephemeris.
We managed to identify \noccs candidates of stellar occultations between the period of January, 2016 and December, 2020. We made observational tests for the prediction of the event of Himalia in March 03, 2015. The stars and objects were observed close to the date predicted and in the same field to minimize errors and the obtained relative satellite-star positions were used to evaluate the predictions. The comparisons between the predictions and the observation tests show a good agreement. We call attention for an occultation by Triton of a bright star (R=12.5) in October 5, 2017. The event can be observed from Europe and the east coast of USA and will be a good opportunity to access the current state of the atmosphere of the satellite. We note that Jupiter will cross the galactic plane in 2019-2020 and Saturn in 2018. This results in a large number of occultations predicted for that period.
We discuss how the successful observation of a stellar occultation by these objects is quite possible and present some of those potential occultations.
\end{abstract}

@@

We present in this paper stellar occultation predictions for the 8 major irregular satellites of Jupiter (Himalia, Elara, Pasiphae, Lysithea, Carme, Ananke, Sinope and Leda), Phoebe of Saturn and, Nereid and Triton of Neptune. Phoebe, being the most studied object with a good measured size, can be used to calibrate and evaluate the technique for similar objects.

@@

\cite{GomesJunior2015} obtained 6523 suitable positions for 18 irregular satellites between 1992 and 2014 with an estimated error in the positions of about 60 to 80 mas. For some satellites the number of positions obtained is comparable to the number used in the numerical integration of orbits by the JPL \citep{Jacobson2012}. They pointed out that the ephemeris of the irregular satellites have systematic errors that may reach 200 mas for some satellites. For an object at the distance of Jupiter, this represents an error larger than 700 km in the shadow path. Using the positions obtained by \cite{GomesJunior2015} we produced a specific short-time ephemeris for the satellites of Jupiter and Phoebe to better predict stellar occultations by these objects.

@@

In section \ref{Sec: integration} we show the building of the production of the new ephemeris. In section \ref{Sec: predictions}, we present the predictions of the stellar occultations by irregular satellites and how they were made. Some tests made to check the accuracy of the predictions are presented in section \ref{Sec: testes} and the final discussion is presented in section \ref{Sec: discussion}.

@@

A total of \noccs events were identified between January 2016 and December 2020. In Table \ref{Tab: satellite-occultation} we present the number of stellar occultations predicted by year for each satellite. Table \ref{Tab: occ-list} shows a sample of the catalogue of occultations generated and their parameters, which are necessary to produce occultation maps. Since these objects are very small, the duration of each event is a few seconds. All the occultation tables and maps will be publicly available at the CDS. No star brighter than MagR*=18 will be occulted by Triton in 2016. For this satellite, we cut events with stars fainter than R=18, since for Triton the flux drop during such an occultation would be very small. In Fig. \ref{Fig: ocultacao} we show an example of an occultation map. This is one of the only two found occultation by Triton and it will happen in October 05, 2017. This event can be observed from Europe and the east coast of USA and will be a great opportunity to study in high resolution the atmosphere of the satellite.

@@

>>>>

\title[New orbits and predictions of stellar occultations by irregular satellites up to 2020]{New numerical integration of orbits of irregular satellites and prediction of stellar occultations up to 2020}
\author[A. R. Gomes-J\'unior, M. Assafin, L. Beauvalet et al.]{A. R. Gomes-J\'unior$^{1}$\thanks{E-mail: altair08@astro.ufrj.br},
M. Assafin$^{1,\dag}$,
L. Beauvalet$^{2,3}$,
J. Desmars$^{4}$,
R. Vieira-Martins$^{1,2,5,\dag}$,\newauthor
J. I. B. Camargo$^{2,5}$,
B. E. Morgado$^{1,2}$
F. Braga-Ribas$^{2,6}$
\\
$^{1}$Observat\'orio do Valongo/UFRJ, Ladeira Pedro Ant\^onio 43,
CEP 20.080-090 Rio de Janeiro - RJ, Brazil\\
$^{2}$Observat\'orio Nacional/MCTI, R. General Jos\'e Cristino 77, CEP 20921-400 Rio de Janeiro - RJ, Brazil\\
$^{3}$Observatoire de Paris/SYRTE, 77 Avenue Denfert Rochereau 75014 Paris, France\\
$^{4}$Institut de m\'ecanique c\'eleste et de calcul des \'eph\'em\'erides - Observatoire de Paris, UMR 8028 du CNRS, 77 Av. Denfert-Rochereau,\\ 75014 Paris, France\\
$^{5}$Laborat\'orio Interinstitucional de e-Astronomia - LIneA, Rua Gal. Jos\'e Cristino 77, Rio de Janeiro, RJ 20921-400, Brazil\\
$^{6}$Federal University of Technology - Paran\'a (UTFPR / DAFIS), Rua Sete de Setembro, 3165, CEP 80230-901, Curitiba, PR, Brazil
}
\begin{document}
\newcommand{\noccs}{5457 } % total number of occultations

@@

\begin{abstract}
Due to their orbital configurations, it is common belief that the irregular satellites were probably captured by the giant planets in the early solar system. It is important to know their physical parameters, such as size, shape, albedo and composition, to trace back their true origin. The best ground-based technique to determine size and shape, and thus constrain the albedo and in a broader sense composition, is the observation of stellar occultations by these objects.
We aim to predict stellar occultations for the eight largest irregular satellites of Jupiter: Himalia, Elara, Pasiphae, Carme, Lysithea, Sinope, Ananke and Leda, and for the irregular satellites Phoebe of Saturn and Nereid of Neptune, and also for Triton.
We identified candidates to stellar occultations by the irregular satellites from the UCAC4 catalogue and from a catalogue of stars in the sky-path of Neptune obtained from observations made with the ESO2p2/WFI (2.2 m Max-Planck ESO telescope with the Wide Field Imager) instrument. These catalogues were crossed with the ephemeris of the satellites to identify stellar occultations. We used a new ephemeris based solely on the observations from \cite{GomesJunior2015} to generate predictions for the short-time future of the satellites of Jupiter. For Phoebe, we used an updated ephemeris and for Nereid and Triton we used recently published ephemeris.
We managed to identify \noccs candidates of stellar occultations between the period of January, 2016 and December, 2020. We made observational tests for the prediction of the event of Himalia in March 03, 2015. The stars and objects were observed close to the date predicted and in the same field to minimize errors and the obtained relative satellite-star positions were used to evaluate the predictions. The comparisons between the predictions and the observation tests show a good agreement. We call attention for an occultation by Triton of a bright star (R=12.5) in October 5, 2017. The event can be observed from Europe and the east coast of USA and will be a good opportunity to access the current state of the atmosphere of the satellite. We also note that Jupiter will cross the galactic plane in 2019-2020 and Saturn in 2018. This results in a large number of occultations predicted for that period.
We discuss how the successful observation of a stellar occultation by these objects is quite possible and present some of those potential occultations.
\end{abstract}

@@

We present in this paper new numerical integration of the orbits of the 8 major irregular satellites of Jupiter (Himalia, Elara, Pasiphae, Lysithea, Carme, Ananke, Sinope and Leda) for predict stellar occultations. This is a specific short-time ephemeris using only the positions obtained by \cite{GomesJunior2015}. \cite{GomesJunior2015} obtained 6523 suitable positions for 18 irregular satellites between 1992 and 2014 with an estimated error in the positions of about 60 to 80 mas. For some satellites the number of positions obtained is comparable to the number used in the numerical integration of orbits by the JPL \citep{Jacobson2012}. They pointed out that the ephemeris of the irregular satellites have systematic errors that may reach 200 mas for some satellites. For an object at the distance of Jupiter, this represents an error langer than 700 km in the shadow path of an occultation.

For Phoebe, we updated the ephemeris of \cite{Desmars2013} using the observations of \cite{GomesJunior2015}, observations from Minor Planet Center and observations from Flagstaff. For Nereid and Triton we used the most recent ephemeris available.

Also, with new and precise ephemeris, we present stellar occultation predictions for these 11 objects up to 2020. Phoebe, being the most studied object with a good measured size, can be used to calibrate and evaluate the technique for similar objects.

@@

@@

In section \ref{Sec: integration} we show the building of the new ephemeris. In section \ref{Sec: predictions}, we present the predictions of the stellar occultations by irregular satellites and how they were made. Some tests made to check the accuracy of the predictions are presented in section \ref{Sec: testes} and the final discussion is presented in section \ref{Sec: discussion}.

@@

A total of \noccs events were identified between January 2016 and December 2020. In Table \ref{Tab: satellite-occultation} we present the number of stellar occultations predicted by year for each satellite. It is possible to see an increase in the number of events found for Phoebe in 2018 and for the satellites of Jupiter in 2019-2020. This is because at that periods these satellites will cross the apparent galactic plane. In contrast, for Nereid and Triton, almost no occultation was found because Neptune is in a very low dense region of stars and move more slowly than the other planets. We call attention that about 10\% (516) of the events will involve stars brighter than magR=14 (and almost 25\% brighter than magR=15), which helps the attempt of amateur observers.

Table \ref{Tab: occ-list} shows a sample of the catalogue of occultations generated and their parameters, which are necessary to produce occultation maps. Since these objects are very small, the duration of each event is a few seconds. All the occultation tables and maps will be publicly available at the CDS. No star brighter than MagR*=18 will be occulted by Triton in 2016. For this satellite, we cut events with stars fainter than R=18, since for Triton the flux drop during such an occultation would be very small. In Fig. \ref{Fig: ocultacao} we show an example of an occultation map. This is one of the only two found occultation by Triton and it will happen in October 05, 2017. This event can be observed from Europe and the east coast of USA and will be a great opportunity to study in high resolution the atmosphere of the satellite.

@@

For this, we generated new numerical integration of the orbits of the irregular satellites of Jupiter. A specific short-time ephemeris were obtained using only the positions of \cite{GomesJunior2015}. The systematic errors found in the JPL ephemeris \citep{Jacobson2012} were corrected. We also updated the ephemeris of \cite{Desmars2013} using the observations of \cite{GomesJunior2015}, from MPC and from Flagstaff. For Nereid and Triton we used the most recent ephemeris available \citep[, respectively]{Emelyanov2011, Emelyanov2015}.
